### 2013-04-03 Docs/MESA/instructions.tex
update MESA docs

Modules for Experiments in Stellar Astrophysics ({\sf MESA}) is a collection 
of open source, robust and efficient modules developed for a wide range of 
applications in computational stellar astrophysics. The modules include 
macroscopic and microscopic physics such as nuclear reaction rates, opacity, 
equation of state, element diffusion data and atmosphere boundary conditions 
in addition to numerical routines. This document describes how to link the 
{\sf MESA} modules to an outside program such as {\sf MAESTRO}.
\footnote{This
  interface work was done by Ryan Orvedahl---questions (and credit)
  should be directed to him}.

@@

To install {\sf MESA}, download the source code from the {\sf MESA} website, 
{\tt http://mesa.sourceforge.net}. Check the website for the latest version, 
which is 4442 with the writing of this document. The codes and algorithms 
discussed here were created when the current version of {\sf MESA} was 
version 4411. Between versions 4411 and 4442, there are minor changes that 
will cause some of the included routines to crash. To download the source code 
associated with version 4411, use the following command:
\begin{lstlisting}[language=bash,mathescape=false]
  svn co -r 4411 http://mesa.svn.sourceforge.net/svnroot/mesa/trunk mesa
\end{lstlisting}
This will download the source into a directory called {\tt mesa} located 
in the directory where this command was run. This step might take some time 
($\sim$10 min depending on the internet connection) possibly because the 
final {\tt mesa} directory will take up roughly 2.2 GB.

@@

\item {\tt BLAS} and {\tt LAPACK}: Set the location of {\tt LAPACK} and 
{\tt BLAS} provided with the {\sf MESA} source tree using the following 
definitions:
\begin{lstlisting}[language=make,mathescape=false,upquote=true]
  WHICH_LAPACK = USE_SRCS
  LOAD_LAPACK = -lmesalapack
  WHICH_BLAS = USE_SRCS
  LOAD_BLAS = -lmesablas
  MKL_INCLUDE =
\end{lstlisting}

@@

\item {\tt LOAD\_MATRIX}: {\tt LOAD\_MATRIX} is first defined near the end of 
the file around line $\sim$275. The purpose of this variable is to load all 
the libraries associated with linear algebra including {\tt LAPACK}, {\tt BLAS} 
and a few others. The lines we wish to change look like this:
\begin{lstlisting}[language=make,mathescape=false]
  #LOAD_MATRIX = -lmtx -lconst -lmesaklu $(LOAD_LAPACK) $(LOAD_BLAS)
  # for ifort on mac, we need the following def for LOAD_MATRIX 
  # because of loader bugs (as of July 2012)
  LOAD_MATRIX = -lmtx $(MESA_DIR)/const/make/const_def.o \
      $(MESA_DIR)/const/make/const_lib.o -lmesaklu $(LOAD_LAPACK) $(LOAD_BLAS)
  LOAD_MESA_NUMERICS = -linterp_2d -linterp_1d -lnum -lutils -lalert \
      $(LOAD_MATRIX)
\end{lstlisting}
We must uncomment the first {\tt LOAD\_MATRIX} and comment out the second one:
\begin{lstlisting}[language=make,mathescape=false]
  LOAD_MATRIX = -lmtx -lconst -lmesaklu $(LOAD_LAPACK) $(LOAD_BLAS)
  # for ifort on mac, we need the following def for LOAD_MATRIX 
  # because of loader bugs (as of July 2012)
  #LOAD_MATRIX = -lmtx $(MESA_DIR)/const/make/const_def.o \
  #    $(MESA_DIR)/const/make/const_lib.o -lmesaklu $(LOAD_LAPACK) $(LOAD_BLAS)
  LOAD_MESA_NUMERICS = -linterp_2d -linterp_1d -lnum -lutils -lalert \
       $(LOAD_MATRIX)
\end{lstlisting}

\item {\tt install} script: In the {\tt MESA} directory, open the {\tt install} 
script and comment out two lines. We want to change:
\begin{lstlisting}[language=bash,mathescape=false]
  do_one atm

  do_one sample

  do_one star

  do_one adipls
\end{lstlisting}
to this:
\begin{lstlisting}[language=bash,mathescape=false]
  do_one atm

  #do_one sample

  #do_one star

  #do_one adipls
\end{lstlisting}
There are routines in the {\tt star} package and {\tt adipls} that rely on 
the {\tt PGSTAR} modules. We do not wish to use these modules and the easiest 
way around this is to not install both {\tt star} and {\tt adipls} because 
neither one is needed to include {\sf MESA} microphysics modules in 
{\sf MAESTRO}. {\tt star} is the 1D stellar evolution module and {\tt adipls} 
module is used for asteroseismology. The {\tt sample} module is simply a 
sample directory that is not needed for using the main microphysics modules.

\item Edit 1 source file: An undefined reference to 
{\tt omp\_get\_max\_threads} occurs in the {\tt const\_def.f} file located in 
{\tt const/public/const\_def.f}.

In the {\tt const\_def.f} file, around line $\sim$164, in the 
{\tt do\_const\_init} subroutine, there is a line that reads:
\begin{lstlisting}[language=fortran,mathescape=false]
  omp_num_threads = omp_get_max_threads()
\end{lstlisting}
Our first change is to declare a temporary character variable in the 
({\tt do\_const\_init} subroutine):
\begin{lstlisting}[language=fortran,mathescape=false]
  character(len=16) :: temp_max_threads
\end{lstlisting}
Then we comment out the above line ({\tt omp\_num\_threads = ...}) and add the 
following (all in the {\tt do\_const\_init} subroutine):
\begin{lstlisting}[language=fortran,mathescape=false]
  !omp_num_threads = omp_get_max_threads()
  call get_environment_variable("OMP_NUM_THREADS", temp_max_threads)
  read(temp_max_threads, *) omp_num_threads
\end{lstlisting}

@@

Anytime a change is made to a makefile or any source code, the module must 
be recompiled, which is a three step process involving the {\tt clean}, 
{\tt mk} and {\tt export} scripts provided in each module directory. The 
first step is to clean out any old files using the {\tt ./clean} script. 
Step two is to make the module using the {\tt ./mk} script. The last step is 
to export the {\tt .mod} files and the library to the right directories using 
the {\tt ./export} script. Section \ref{sec:compile_scripts} contains a 
simple script that is meant to help with compiling modules. It cleans, makes 
and exports every module and should be run in the top {\tt mesa/} directory.

@@

In {\tt mesa/star/private/micro.f} there are three places 
to set the {\tt use\_omp} flag to False. They occur in the {\tt do\_eos}, 
{\tt do\_neu} and {\tt do\_net} routines. The fourth place to set the 
{\tt use\_omp} flag to False is contained in the {\tt do\_kap} routine in 
{\tt mesa/star/private/opacities.f}.

>>>>

Modules for Experiments in Stellar Astrophysics ({\sf MESA}) is a collection 
of open source, robust and efficient modules developed for a wide range of 
applications in computational stellar astrophysics. The modules include 
macroscopic and microscopic physics such as nuclear reaction rates, opacity, 
equation of state, element diffusion data and atmosphere boundary conditions 
in addition to numerical routines. This document describes how to link the 
{\sf MESA} modules to an outside program such as {\sf MAESTRO}\footnote{This
  interface work was done by Ryan Orvedahl---questions (and credit)
  should be directed to him}.

@@

To install {\sf MESA}, download the source code from the {\sf MESA} website, 
{\tt http://mesa.sourceforge.net}. Check the website for the latest version, 
which is 4849 when the last upgrades to this document were added. The codes 
and algorithms 
discussed here were created when the current version of {\sf MESA} was 
version 4088 and have been modified accordingly to keep up with the quickly 
channging source code. It might be possible to use older versions of 
{\sf MESA} with 
the information in this document, although there are minor changes and 
upgrades between {\sf MESA} versions that may cause the routines and algorithms 
discussed here to crash. To download the source code associated with version 
4849, issue the following command:
\begin{lstlisting}[language=bash,mathescape=false]
  svn co -r 4849 svn://svn.code.sf.net/p/mesa/code/trunk mesa
\end{lstlisting}
This will download the source into a directory called {\tt mesa} located 
in the directory where this command was run. This step might take some time 
($\sim$3 min depending on the internet connection) possibly because the 
final {\tt mesa} directory will take up roughly 2.1 GB.

@@

\item {\tt BLAS} and {\tt LAPACK}: Set the location of {\tt LAPACK} and 
{\tt BLAS} provided with the {\sf MESA} source tree using the following 
definitions (these should be the default settings):
\begin{lstlisting}[language=make,mathescape=false,upquote=true]
  WHICH_LAPACK = USE_SRCS
  LOAD_LAPACK = -lmesalapack
  WHICH_BLAS = USE_SRCS
  LOAD_BLAS = -lmesablas
  MKL_INCLUDE =
\end{lstlisting}

@@

%============================================================================
% This next \item is unnecessary for version 4849 (it was necessary for 4411)
%============================================================================
%
%\item {\tt LOAD\_MATRIX}: {\tt LOAD\_MATRIX} is first defined near the end of 
%the file around line $\sim$280. The purpose of this variable is to load all 
%the libraries associated with linear algebra including {\tt LAPACK}, 
%{\tt BLAS} and a few others. The lines we wish to change look like this:
%\begin{lstlisting}[language=make,mathescape=false]
%  #LOAD_MATRIX = -lmtx -lconst -lmesaklu $(LOAD_LAPACK) $(LOAD_BLAS)
%  # for ifort on mac, we need the following def for LOAD_MATRIX 
%  # because of loader bugs (as of July 2012)
%  LOAD_MATRIX = -lmtx $(MESA_DIR)/const/make/const_def.o \
%      $(MESA_DIR)/const/make/const_lib.o -lmesaklu $(LOAD_LAPACK) $(LOAD_BLAS)
%  LOAD_MESA_NUMERICS = -linterp_2d -linterp_1d -lnum -lutils -lalert \
%      $(LOAD_MATRIX)
%\end{lstlisting}
%We must uncomment the first {\tt LOAD\_MATRIX} and comment out the second one:
%\begin{lstlisting}[language=make,mathescape=false]
%  LOAD_MATRIX = -lmtx -lconst -lmesaklu $(LOAD_LAPACK) $(LOAD_BLAS)
%  # for ifort on mac, we need the following def for LOAD_MATRIX 
%  # because of loader bugs (as of July 2012)
%  #LOAD_MATRIX = -lmtx $(MESA_DIR)/const/make/const_def.o \
%  #    $(MESA_DIR)/const/make/const_lib.o -lmesaklu $(LOAD_LAPACK) $(LOAD_BLAS)
%  LOAD_MESA_NUMERICS = -linterp_2d -linterp_1d -lnum -lutils -lalert \
%       $(LOAD_MATRIX)
%\end{lstlisting}

%============================================================================
% This next \item is unnecessary for version 4849 (it was necessary for 4411)
%============================================================================
%
%\item {\tt install} script: In the {\tt MESA} directory, open the 
%{\tt install} script and comment out two lines. We want to change:
%\begin{lstlisting}[language=bash,mathescape=false]
%  do_one atm
%
%  do_one sample
%
%  do_one star
%
%  do_one adipls
%\end{lstlisting}
%to this:
%\begin{lstlisting}[language=bash,mathescape=false]
%  do_one atm
%
%  #do_one sample
%
%  #do_one star
%
%  #do_one adipls
%\end{lstlisting}
%There are routines in the {\tt star} package and {\tt adipls} that rely on 
%the {\tt PGSTAR} modules. We do not wish to use these modules and the easiest 
%way around this is to not install both {\tt star} and {\tt adipls} because 
%neither one is needed to include {\sf MESA} microphysics modules in 
%{\sf MAESTRO}. {\tt star} is the 1D stellar evolution module and {\tt adipls} 
%module is used for asteroseismology. The {\tt sample} module is simply a 
%sample directory that is not needed for using the main microphysics modules.

%============================================================================
% This next \item is necessary when compiling in serial
%============================================================================
%
\item Edit 1 source file: An undefined reference to 
{\tt omp\_get\_max\_threads} occurs in the {\tt const\_def.f} file located in 
{\tt const/public/const\_def.f} when compiling in serial.

In the {\tt const\_def.f} file, around line $\sim$168, in the 
{\tt do\_const\_init} subroutine, there is a line that reads:
\begin{lstlisting}[language=fortran,mathescape=false]
  omp_num_threads = omp_get_max_threads()
\end{lstlisting}
Our first change is to declare a temporary character variable in the 
({\tt do\_const\_init} subroutine):
\begin{lstlisting}[language=fortran,mathescape=false]
  character(len=16) :: temp_max_threads
\end{lstlisting}
Then we comment out the above line ({\tt omp\_num\_threads = ...}) and add the 
following (all in the {\tt do\_const\_init} subroutine):
\begin{lstlisting}[language=fortran,mathescape=false]
  !omp_num_threads = omp_get_max_threads()
  call get_environment_variable("OMP_NUM_THREADS", temp_max_threads)
  read(temp_max_threads, *) omp_num_threads
\end{lstlisting}

@@

Anytime a change is made to a makefile or any source code, the module must 
be recompiled, which is a three step process involving the {\tt clean}, 
{\tt mk} and {\tt export} scripts provided in each module directory. The 
first step is to clean out any old files using the {\tt ./clean} script. 
Step two is to make the module using the {\tt ./mk} script. The last step is 
to export the {\tt .mod} files and the library to the right directories using 
the {\tt ./export} script. Section \ref{sec:compile_scripts} discusses a few 
simple scripts that are meant to help with compiling modules.

@@

To find where all to set the {\tt use\_omp} flag to False, execute the 
following command:
\begin{lstlisting}[language=bash,mathescape=false]
  find $(MESA_DIR) -name "*.f" -exec grep -Hn --color use_omp {} \;
\end{lstlisting}
This will return fortran files located in the {\tt mesa/star/private} 
directory.
